### 2018-07-17 msreview.tex
Backup

\documentclass[11pt, a4paper]{report}
% FOR COLORFUL TABLES
\usepackage{booktabs}
\usepackage[table]{xcolor}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{pdfpages}
\usepackage{array} % to make table column italic
\usepackage{textcomp} % to support the trademark symbol
\usepackage{minted}

\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
\usepackage{wallpaper}
\usepackage{palatino}
\usepackage{changepage}
\usepackage{mhchem} % for isotope nomeclature support

% Setup captions
%\captionstyle[\centering]{\centering}
%\changecaptionwidth
%\captionwidth{0.8\linewidth}

% Protect against widows and orphans
%\clubpenalty=10000
%\widowpenalty=10000

%\linespread{1.2}

%\raggedbottom

%\chapterstyle{ger}

%\maxsecnumdepth{subsection}

%%  Setup fancy style quotation
%%  ==================================================================
%\usepackage{tikz}
%\usepackage{framed}

%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font

% Make commands for the quotes
%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}

% select a colour for the shading
%\definecolor{shadecolor}{rgb}{1,1,1}

% wrap everything in its own environment
%\newenvironment{shadequote}% 
%{\begin{snugshade}\begin{quote}\openquote}
%{\hfill\closequote\end{quote}\end{snugshade}}

%% MY STUFF

%% Language and font encodings

%% Sets page size and margins
%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{font={small},skip=0.25\baselineskip}
\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
\graphicspath{{./plots/}{./figures/}}

%\fancypagestyle{title}{%
%  \setlength{\headheight}{120pt}%
%  \fancyhf{}% No header/footer
%  \renewcommand{\headrulewidth}{0pt}% No header rule
%  \renewcommand{\footrulewidth}{0pt}% No footer rule
%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
%}%

%\usepackage{amsmath}
%\usepackage[format=plain]{caption}
\usepackage{subcaption}
\usepackage{float} % force float
\usepackage[nottoc]{tocbibind}
%\usepackage[margin=-50pt]{caption}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{cleveref} % for enumerate/itemize item referencing

%\setlength{\parindent}{4em}
%\setlength{\parskip}{1em}
%\usepackage{mathpazo} % Use the Palatino font
\usepackage{parskip}% http://ctan.org/pkg/parskip
\usepackage{xspace}
\usepackage{wrapfig}
\usepackage[export]{adjustbox}
\usepackage{textcomp}
\usepackage[bottom]{footmisc}

\usepackage{titling}
\newcommand{\subtitle}[1]{%
  \posttitle{%
    \par\end{center}
    \begin{center}\large#1\end{center}
    \vskip0.5em}%
}

\renewcommand\thesubfigure{\arabic{subfigure}}  
\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}

\usepackage{listings}
\lstset{basicstyle=\ttfamily,
  showstringspaces=false,
  commentstyle=\color{red},
  keywordstyle=\color{blue}
}

\makeatletter
\renewcommand\p@subfigure{\thefigure\,}
\renewcommand\thesubfigure{\alph{subfigure})}
% If desired for table as well:
% \renewcommand\p@subtable{\thetable\,}
% \renewcommand\thesubtable{\alph{subtable})}
\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
\makeatother

\renewcommand{\baselinestretch}{1.5}
\DeclareCaptionSubType*[arabic]{figure}

%\makeatletter
%\renewcommand\thesection{}
%\renewcommand\thesubsection{\@arabic\c@subsection}
%\makeatother

\usepackage{mathtools}
\usepackage{sectsty}

\chapternumberfont{\LARGE} 
\chaptertitlefont{\LARGE}

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%

% Swap the definition of \abs* and \norm*, so that \abs
% and \norm resizes the size of the brackets, and the 
% starred version does not.
\makeatletter
\let\oldabs\abs
\def\abs{\@ifstar{\oldabs}{\oldabs*}}
%
\let\oldnorm\norm
\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
\makeatother

\usepackage{setspace}
\doublespacing

%\pagestyle{plain} % default for report

\usepackage{etoolbox}
\makeatletter
\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
\makeatother

\usepackage{acro}
\acsetup{first-style=short}

\DeclareAcronym{MS}{
  short = MS ,
  long  = Mass Spectrometry,
  class = abbrev
}

\DeclareAcronym{MS1}{
  short = MS1 ,
  long  = first tandem MS analyzer ,
  class = abbrev
}

\DeclareAcronym{MS2}{
  short = MS2 ,
  long  = second tandem MS analyzer ,
  class = abbrev
}

\DeclareAcronym{MS/MS}{
  short = MS/MS ,
  long  = Tandem mass spectrometry,
  class = abbrev
}

\DeclareAcronym{m/z}{
  short = m/z ,
  long  = mass charge ratio,
  class = abbrev
}

\DeclareAcronym{MALDI}{
  short = MALDI ,
  long  = Matrix-Assisted Laser-Desorption Ionization,
  class = abbrev
}

\DeclareAcronym{ESI}{
  short = ESI ,
  long  = Electrospray Ionization,
  class = abbrev
}

\DeclareAcronym{TOF}{
  short = TOF ,
  long  = Time of Flight,
  class = abbrev
}

\DeclareAcronym{IT}{
  short = IT ,
  long  = Ion Trap,
  class = abbrev
}

\DeclareAcronym{Q}{
  short = Q,
  long  = Quadrupole,
  class = abbrev
}

\DeclareAcronym{FT-ICR}{
  short = FT-ICR ,
  long  = Fourier Transform Ion Cyclotron Resonance,
  class = abbrev
}

\DeclareAcronym{HPLC}{
  short = HPLC ,
  long  = High Pressure Liquid Chromatography,
  class = abbrev
}

\DeclareAcronym{MCMC}{
  short = MCMC,
  long  = Markov Chain Monte Carlo,
  class = abbrev
}

\DeclareAcronym{PTM}{
  short = PTM ,
  long  = Post-Translational Modification,
  class = abbrev
}

\DeclareAcronym{DAG}{
  short = DAG ,
  long  = Directed Acyclic Graph,
  class = abbrev
}

\DeclareAcronym{NZ}{
  short = NZ ,
  long  = Novozymes A/S,
  class = abbrev
}

\DeclareAcronym{FDR}{
  short = FDR ,
  long  = False Discovery Rate,
  class = abbrev
}

\DeclareAcronym{RT}{
  short = RT ,
  long  = Retention time,
  class = abbrev
}

\DeclareAcronym{log2FC}{
  short = log2FC ,
  long  = $\log_2$(FC),
  class = abbrev
}

\DeclareAcronym{MBR}{
  short = MBR ,
  long  = Match Between Runs,
  class = abbrev
}

\DeclareAcronym{ADVI}{
  short = ADVI ,
  long  = Automatic Differentiation Variational Inference,
  class = abbrev
}

\DeclareAcronym{NUTS}{
  short = NUTS ,
  long  = No-U-Turn-Sampler,
  class = abbrev
}

\DeclareAcronym{MGF}{
  short = MGF ,
  long  = Mascot Generic Format,
  class = abbrev
}

\DeclareAcronym{XIC}{
  short = XIC ,
  long  = Extracted Ion Current,
  class = abbrev
}

\DeclareAcronym{SC}{
  short = SC ,
  long  = Spectral Counting,
  class = abbrev
}

\DeclareAcronym{TPP}{
  short = TPP ,
  long  = Trans Proteomic Pipeline,
  class = abbrev
}

\DeclareAcronym{NHST}{
  short = NHST ,
  long  = Null-Hypothesis Significance Testing,
  class = abbrev
}

%%  Begin document
%%  ==================================================================
\begin{document}

%%  Begin title page
%%  ==================================================================
    \thispagestyle{empty}
    \ULCornerWallPaper{1}{nat-farve.pdf}
    \ULCornerWallPaper{1}{nat-en.pdf}
    \begin{adjustwidth}{-3cm}{-1.5cm}
    \vspace*{-1cm}
    \textbf{\Huge Bioinformatics Master Thesis} \\
    \vspace*{2.5cm} \\
    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
    \vspace*{.1cm} \\
    %{\huge No subtitle is available} \\
    \begin{tabbing}
    % adjust the hspace below for the longest author name
    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
    %Second Author \> \texttt{<second@author.com>} \\       
    \\[10.5cm]

    \textbf{\Large Supervisors} \\
    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\

    \end{tabbing}
    \end{adjustwidth}
    \ClearWallPaper
%%  ==================================================================
%%  End title page

%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
\setcounter{tocdepth}{2}
\tableofcontents % Indholdsfortegnelse
%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

\thispagestyle{empty}
\printacronyms[include-classes=abbrev,name=Abbreviations]

\newpage
\setcounter{page}{1}

% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
\let\oldthebibliography\thebibliography
\let\endoldthebibliography\endthebibliography
\renewenvironment{thebibliography}[1]{
  \begin{oldthebibliography}{#1}
    \setlength{\itemsep}{0em}
    \setlength{\parskip}{0em}
}
{
  \end{oldthebibliography}
}

\section*{Preface}

TODO: get nomenclature straight: what is a sample and what is a fraction

\chapter*{Introduction}
\label{chap:introduction}

\section{Aminoacids and proteins}

Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.

Proteins are made up of 20 basic units, called aminoacids. All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid. The side chain differs between aminaocids and generates them from each other. A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid. Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
  \caption{CAPTION AND REFERENCE}
  \label{fig:aminoacids}
\end{figure}

Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them. Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}. The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.

While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite. One frequent instance of such modifications is the oxidation of methionine.

\section{The protein-focused biotechnology industry}

Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks. The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans. This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).

\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes. The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact. Notorious examples of such applications include waste-water treatment, household care and the baking industry.

The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors. The refinement of the currently used tools and the development of new ones could be of great significance for the company.

Protein research can be approached from different angles. This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.

\section{Objectives of the Thesis}
\label{sec:objectives}

In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:

\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.

\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.

\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.

\section{Structure of the Thesis}

An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}. The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}. Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.

@@

On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance. A wide range of algorithms are available to process these data and output estimates of protein abundance. All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements. They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.

@@

\chapter{A label-free quantification proteomics pipeline}
\label{chap:pipeline}

\section*{Summary}

A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016} , was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software. The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) \ac{MBR} and feature extraction, (IV) protein inference and (V) relative quantification. Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.  A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}. The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.

\section{Introduction}

\subsection{Background}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{pipeline}
\caption{Schema of the presented pipeline. First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}). After that, spectra are searched by means of a search engine to perform the PSM step. Thereafter, PSMs are validated and the most likely set proteins is inferred. If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed. The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
\label{fig:pipeline}
\end{figure}

Several proteomics pipelines are available on the internet under different licensing conditions. Many are released as closed-source commercial software, where information on how the program works is kept from the user. This is a serious drawback as it hinders the study of the implemented models and its customisation. Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability. MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.

\subsection{Goals}

The development of a pipeline attempting to achieve the following goals will be described in this chapter.

\item Fully command line, documented and Linux-supported interface for easy automation and scalability.
\item Open-source for easy customisation and innovation of analyses.
\item Free-licensed and cost-free, so anybody with the knowledge can run it, democratizing the analyses.
\end{enumerate}

\section{Materials and Methods}

\subsection{Data generation and loading}

The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. Briefly, the \textit{Homo sapiens} and \textit{E. coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination. Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions. This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files. One file was missing in the repository. The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format.

The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file. The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.

%\href{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}.

\subsection{Decoy database preparation and search}
\label{subsec:database_preparation}

The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine \cite{Kim2014} by means of the SearchGUI \texttt{SearchCLI} tool \cite{Barsnes2018} utility. The search parameters were set using the \texttt{IdentificationParametersCLI}. In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q. Moreover, C carbamidomethylation was set as fixed modification. The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide. Up to two missed cleavages were allowed. The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.

The target database was created by combining the Uniprot proteomes for \textit{E. coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018. The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.

\subsection{Quality control and validation}

The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}. By default, the FDR was set to 1\%. PEP and confidence statistics were computed using the PeptideShaker built-in algorithms. Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.

\subsection{Data refinement}

The moFF command line utility \cite{Argentini2016} was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster. This required passing the original RAW files, together with the Default PSM report from PeptideShaker. Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.

\subsection{Quantification}

Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF. Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function. In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.

Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually. The peptide and fraction effects were considered random, while the condition was treated as a fixed effect. The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function. Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.

Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file. Intensities were log2 transformed and averaged. The difference between conditions H and L was taken as estimate of the log2FC. Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R. P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.

\subsection{Code implementation}

\section{Results}

\subsection{Thousands of spectra were identified and validated}

The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra. PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation. As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values. Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter. The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.

\begin{figure}[!h]
\centering
\begin{subfigure}{.45\textwidth}
  \centering
  \caption*{A}
  \includegraphics[width=.9\linewidth]{target_vs_decoy}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
  \centering
    \caption*{B}
  \includegraphics[width=.9\linewidth]{score_vs_confidence}
\end{subfigure}
\bigskip

\begin{subfigure}{.45\textwidth}
  \centering
    \caption*{C}
  \includegraphics[width=.95\linewidth]{qc}
\end{subfigure}
\begin{subfigure}{.45\textwidth}
  \centering
    \caption*{D}
  \includegraphics[width=.95\linewidth]{roc}
\end{subfigure}
\caption{Quality control and validation of the 13th fraction of the third replicate in condition L. \textbf{A} Score distribution for matches to the decoy and the target databases. \textbf{B} Evolution of the PSM score with confidence. The implemented cutoff at FDR of 1 \% is displayed with a red vertical line. \textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status. \textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search. The cutoff is displayed as a red dot.}
\label{figure:qc_validation}
\end{figure}

 The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.

The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples. However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).

\begin{figure}[!h]
\includegraphics[width=0.8\textwidth]{match_percent}
\caption{Percentage of matched spectra in all samples. The total number of spectra per sample ranged between 5894 and 20249. \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa. \textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation. The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
\label{fig:match_percent}
\end{figure}

\subsection{Protein inference}

\begin{table}[H]
\begin{tabular}{ccccc}
  \toprule
 Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
  \midrule
 46595 & 27384 & 19211 & 25056 & 10039 \\
 \bottomrule
\end{tabular}
\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
\label{tab:protein_inference}
\end{table}

A total of 46595 peptides were inferred to be present in at least one of the fractions (see table \ref{tab:protein_inference}). Of them more than 27 thousand mapped to a unique protein. However, a significant amount mapped to more than one protein. Such peptides are called non-unique peptides. The 25056 detected proteins were grouped into 10039 protein groups. Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them. Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from. Protein grouping decreased the amount of  considered protein entities to 10039.

Protein groups are very frequent due to several factors. For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve. PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.

However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.

\subsection{MBR and apex intensity extraction quantification}

The match between runs step allowed for increased identifications by transferring successful matches between replicate runs. The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{mbr_combined}
\caption{Match Between Runs with moFF. \textbf{A} Count of identifications on each run segreagated by source. More than 4k spectra were identified and validated by SearchGUI+peptideShaker. In this particular case, hundreds of new identifications were accomplished. \textbf{B} Match visualization. Every dot represents a peptide shared across 2 runs. The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis. The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
\label{fig:mbr}
\end{figure}

Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).

\begin{figure}[!h]
\begin{subfigure}{.9\textwidth}
  \centering
    \caption*{A}
\includegraphics[width=.9\linewidth]{peptide_profile}
\end{subfigure}
\bigskip

\begin{subfigure}{.9\textwidth}
  \centering
    \caption*{B}
\includegraphics[width=.9\linewidth]{density_ratio}
\end{subfigure}
\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E. coli}. The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E. coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs. \textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E. coli} data in condition H was observed, confirming an acceptable performance of the protocol.}
\label{fig:apex_intensity}
\end{figure}

\subsection{Quantification}
\label{subsec:quantification}

The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L. During quantification, it is important to take into account as many of the effects influencing protein quantification. In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished. The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed. Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.

The result of the process was the successful quantification of 5307 protein groups out of 10039 (see table \ref{tab:quantification_table}).

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Sun Jul 15 18:17:23 2018
\begin{table}[ht]
\centering
\begin{tabular}{llll>{\itshape}l}
  \toprule
 & \ac{log2FC} & qval & Protein & Organism \\ 
  \midrule
1 & -4.76E-01 & 1.42E-24 & P78527 & Homo sapiens \\ 
   \rowcolor[gray]{0.95}2 & 6.80E-01 & 6.75E-24 & P0A8V2 & E. coli \\ 
  3 & 8.50E-01 & 4.39E-22 & P25516 & E. coli \\ 
   \rowcolor[gray]{0.95}4 & 7.07E-01 & 2.05E-18 & P63284 & E. coli \\ 
  5 & 1.04E+00 & 1.21E-17 & P37095 & E. coli \\ 
   \rowcolor[gray]{0.95}6 & 7.78E-01 & 1.51E-16 & P13029 & E. coli \\ 
  7 & 8.89E-01 & 5.62E-16 & P77804 & E. coli \\ 
   \rowcolor[gray]{0.95}8 & 8.53E-01 & 1.72E-15 & P23721 & E. coli \\ 
  9 & -6.97E-01 & 1.41E-14 & P09874 & Homo sapiens \\ 
   \rowcolor[gray]{0.95}10 & 1.37E+00 & 1.41E-14 & P37666 & E. coli \\ 
   \bottomrule
\end{tabular}
\caption{Results of the quantification pipeline. The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown. Notably, eight correspond to proteins belonging to \textit{E.coli}. The two human proteins exhibited in both cases an absolute value of the \ac{log2FC} below 1.}
\label{tab:quantification_table}
\end{table}

The performance of the quantification can be evaluated thanks to the experimental design of the dataset. Since the \textit{E. coli} proteome was mixed on a 3:1 ratio in condition H, a \ac{log2FC} of $log2(3)=1.58$ is expected for proteins coming from this organism. Likewise, human proteins should have a \ac{log2FC} of 0. As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{curves_plot}
\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available. \textbf{Comp+Rob} is the pipeline presented in the previous sections. \textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine. Finally, MQ+Rob blends MaxQuant upstream analysis with the MSqRob quantification engine. \textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively. \textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
\label{fig:roc_curves}
\end{figure}

Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable. Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.

In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes. Thus, \textit{E. coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives. Moreover, a filter based on the \ac{log2FC} can be applied to enrich the data in positives and facilitate the classification task.
In order to compare the results of the here presented pipeline (Comp+Rob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps. The pipeline combining MaxQuant and MaxLFQ (MQ+LFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+Rob) used MaxQuant up to the fraction normalization step.

The result of this analysis is shown in figure \ref{fig:roc_curves}. Remarkably, Comp+Rob achieved similar results to those resulting from MQ+LFQ. Nonetheless, the latter performed clearly better when the \ac{log2FC} filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant. In all cases MQ+Rob was as good as MQ+LFQ, if not much better.

\begin{figure}[!h]
\centering
\begin{subfigure}{.40\textwidth}
\caption*{A}
\includegraphics[width=.9\textwidth]{vennDiagram}
\end{subfigure}
\begin{subfigure}{.52\textwidth}
\caption*{B}
\includegraphics[width=.95\textwidth]{correlation_plot}
\end{subfigure}
\caption{Agreement between the different pipelines. \textbf{A} The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline. \textbf{B} The log2FC estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
\label{fig:venn_cor}
\end{figure}

>>>>

@@

On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance. A wide range of algorithms are available to process these data and output estimates of protein abundance. All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the distributions symmetrical and thus fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements. They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.

@@
